Countering language attrition with PanLex and the Web of Data

### 2014-02-18 panlex-rdf-paper/2012/MLOD_PanLex/paper_v4.tex
replaced approver with source and shrunk occurences table

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{images/schema-new.png}
  \caption{The PanLex database schema}
  \label{fig:db-schema}
\end{figure}

@@

An overview of the number of instances per entity as of January 2014 is given in~\autoref{fig:plx-entity-counts}.
\begin{table}
  \centering\begin{scriptsize}
  \begin{tabular}{lrclr}
    Entity             & Instances  \\
    \cline{1-2}
    Denotations        & 54,278,860 \\  % SELECT COUNT(dn) FROM dn;
    Languages          &      7,843 \\ % SELECT COUNT(lc) FROM lc; -- all language codes (ISO 639-3, ISO 639-2, ISO 639-5) known in the PanLex db
    Language Varieties identified           &      9,310 \\  % SELECT COUNT(lv) FROM lv;
    Language Varieties with data           &      9,239 \\  % 
    Meanings           & 20,773,371 \\  % SELECT COUNT(mn) FROM mn;
    Sources being consulted         &      4,190 \\  % 
    Sources already consulted         &      1,453 \\  % 
    Expressions        & 19,790,453 \\ % SELECT COUNT(ex) FROM ex;
    Definitions        &  2,747,892 \\ % SELECT COUNT(df) FROM df;
    Users              &          23 \\ % SELECT COUNT(us) FROM us;
  \end{tabular}
  \end{scriptsize}
  \caption{Number of instances of main entities in the PanLex database}
  \label{fig:plx-entity-counts}
\end{table}

@@

\begin{itemize}
  \item Expressions are modeled as instances of the class \texttt{\small plx:Expression}.
    Their original and degraded textual representations become the values of the properties \texttt{\small rdfs:label} and \texttt{\small plx:degradedText}, respectively.
    Their corresponding language variety is stated using \texttt{\small plx:languageVariety}.
  \item For language and language varieties the classes \texttt{\small plx:Language} and \texttt{\small plx:LanguageVariety} are introduced.
    \emph{ISO 639-1} and \emph{ISO 639-3 codes} become instances of the classes \texttt{\small plx:Iso639-1Code} and \texttt{\small plx:Iso639-3Code}.
  \item The RDF analog of the PanLex \emph{meaning} is the \texttt{\small plx:Meaning}.
    Entities of this class may have an identifier assigned with the \texttt{\small plx:identifier} property pointing to an \texttt{\small xsd:string} literal.
    Meanings may also have \emph{definitions}, entities of the \texttt{\small plx:Definition} class, giving a textual representation (\texttt{\small rdfs:label}) in a certain language variety (\texttt{\small plx:languageVariety}).
  \item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
    These are entities of the \texttt{\small plx:Denotation} class pointing to meanings and expressions via the properties \texttt{\small plx:denotationMeaning} and \\ \texttt{\small plx:denotationExpression}.
    Denotations may also have a word class assigned to them.
    This can be achieved with the denotation's \texttt{\small plx:wordClass} property pointing to a \texttt{\small plx:WordClass} entity.
  \item All sources share the \texttt{\small plx:Approver} class.
    The characteristics of a source are described using mainly triples with literal objects.
    These are for example \texttt{\small dc:title} to assign the title of a source, \texttt{\small dc:creator} to give an \texttt{\small xsd:string} containing the author's name. At present, we support the different license categories recognized in the database by creating resources of the \texttt{\small plx:License} class.
\end{itemize}

\begin{table}
  \begin{tiny}
  \begin{tabular}{p{52px}p{140px}}
    Class                 & Properties \\
    \toprule
    \texttt{plx:Approver} & \mbox{\texttt{plx:registrationDate}, \texttt{rdfs:label}, \texttt{dc:title},}
                            \mbox{\texttt{dc:creator}, \texttt{plx:license}, \texttt{dc:date}, \texttt{plx:quality},}
                            \mbox{\texttt{foaf:homepage}, \texttt{dc:publisher}, \texttt{dbpedia-owl:isbn}} \\
    \midrule
    \texttt{plx:Language} & \texttt{plx:iso639-3Code}, \texttt{plx:iso639-1Code} \\
    \midrule
    \texttt{plx:LanguageVariety}
                          & \texttt{plx:languageVarietyOf}, \texttt{rdfs:label} \\
    \midrule
    \texttt{plx:Iso639-1Code} & \\
    \midrule
    \texttt{plx:Iso639-3Code} & \\
    \midrule
    \texttt{plx:Expression}
                          & \mbox{\texttt{plx:languageVariety}, \texttt{plx:degradedText},} \texttt{rdfs:label} \\
    \midrule
    \texttt{plx:Meaning}
                          & \mbox{\texttt{plx:approver}, \texttt{plx:identifier},} \texttt{plx:meaningDefinition} \\
    \midrule
    \texttt{plx:Definition}
                          & \texttt{plx:languageVariety}, \texttt{rdfs:label} \\
    \midrule
    \texttt{plx:Denotation}
                          & \texttt{plx:denotationMeaning}, \texttt{plx:denotationExpression}, \texttt{plx:wordClass} \\
    \midrule
    \texttt{plx:WordClass}  
                          & \texttt{rdfs:label} \\
    \midrule
    \texttt{plx:License}  & \texttt{rdfs:label} \\
    \bottomrule
  \end{tabular}
  \end{tiny}
  \caption{Classes and properties used in the PanLex RDF vocabulary. Note that all \texttt{rdf:type} properties are omitted for brevity.}
  \label{tbl:vocabulary}
\end{table}

>>>>

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{images/schema_new.png}
  \caption{The PanLex database schema}
  \label{fig:db-schema}
\end{figure}

@@

An overview of the number of instances per entity as of January 2014 is given in~\autoref{fig:plx-entity-counts}.
%\begin{table}
%  \centering\begin{scriptsize}
%  \begin{tabular}{lrclr}
%    Entity             & Instances  \\
%    \cline{1-2}
%    Denotations        & 54,278,860 \\  % SELECT COUNT(dn) FROM dn;
%    Languages          &      7,843 \\ % SELECT COUNT(lc) FROM lc; -- all language codes (ISO 639-3, ISO 639-2, ISO 639-5) known in the PanLex db
%    Language Varieties identified           &      9,310 \\  % SELECT COUNT(lv) FROM lv;
%    Language Varieties with data           &      9,239 \\  % 
%    Meanings           & 20,773,371 \\  % SELECT COUNT(mn) FROM mn;
%    Sources being consulted         &      4,190 \\  % 
%    Sources already consulted         &      1,453 \\  % 
%    Expressions        & 19,790,453 \\ % SELECT COUNT(ex) FROM ex;
%    Definitions        &  2,747,892 \\ % SELECT COUNT(df) FROM df;
%    Users              &          23 \\ % SELECT COUNT(us) FROM us;
%  \end{tabular}
%  \end{scriptsize}
%  \caption{Number of instances of main entities in the PanLex database}
%  \label{fig:plx-entity-counts}
%\end{table}

%    Users              &         23 \\ % SELECT COUNT(us) FROM us;
\begin{table}
  \centering\begin{scriptsize}
  \begin{tabular}{lrclr}
    Entity             & Instances  & & Entity    & Instances \\
    \cline{1-2}\cline{4-5}
    Denotations        & 54,278,860 & & Languages & 7,843     \\
    Meanings           & 20,773,371 & & Sources   & 4,190     \\
    Expressions        & 19,790,453 & & Users     &    23     \\
    Definitions        &  2,747,892 & & Licenses  &    11     \\
    Language Varieties &      9,310 & & & \\
  \end{tabular}
  \end{scriptsize}
  \caption{Number of instances of main entities in the PanLex database}
  \label{fig:plx-entity-counts}
\end{table}\todo{new license `75 | pl | pup'; What does it mean?}

@@

\begin{itemize}
  \item Expressions are modeled as instances of the class \texttt{\small plx:Expression}.
    Their original and degraded textual representations become the values of the properties \texttt{\small rdfs:label} and \texttt{\small plx:degradedText}, respectively.
    Their corresponding language variety is stated using \texttt{\small plx:languageVariety}.
  \item For language and language varieties the classes \texttt{\small plx:Language} and \texttt{\small plx:LanguageVariety} are introduced.
    \emph{ISO 639-1} and \emph{ISO 639-3 codes} become instances of the classes \texttt{\small plx:Iso639-1Code} and \texttt{\small plx:Iso639-3Code}.
  \item The RDF analog of the PanLex \emph{meaning} is the \texttt{\small plx:Meaning}.
    Entities of this class may have an identifier assigned with the \texttt{\small plx:identifier} property pointing to an \texttt{\small xsd:string} literal.
    Meanings may also have \emph{definitions}, entities of the \texttt{\small plx:Definition} class, giving a textual representation (\texttt{\small rdfs:label}) in a certain language variety (\texttt{\small plx:languageVariety}).
  \item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
    These are entities of the \texttt{\small plx:Denotation} class pointing to meanings and expressions via the properties \texttt{\small plx:denotationMeaning} and \\ \texttt{\small plx:denotationExpression}.
    Denotations may also have a word class assigned to them.
    This can be achieved with the denotation's \texttt{\small plx:wordClass} property pointing to a \texttt{\small plx:WordClass} entity.
  \item All sources share the \texttt{\small plx:Source} class.
    The characteristics of a source are described using mainly triples with literal objects.
    These are for example \texttt{\small dc:title} to assign the title of a source, \texttt{\small dc:creator} to give an \texttt{\small xsd:string} containing the author's name. At present, we support the different license categories recognized in the database by creating resources of the \texttt{\small plx:License} class.
\end{itemize}

\begin{table}
  \begin{tiny}
  \begin{tabular}{p{52px}p{140px}}
    Class                 & Properties \\
    \toprule
    \texttt{plx:Source} & \mbox{\texttt{plx:registrationDate}, \texttt{rdfs:label}, \texttt{dc:title},}
                            \mbox{\texttt{dc:creator}, \texttt{plx:license}, \texttt{dc:date}, \texttt{plx:quality},}
                            \mbox{\texttt{foaf:homepage}, \texttt{dc:publisher}, \texttt{dbpedia-owl:isbn}} \\
    \midrule
    \texttt{plx:Language} & \texttt{plx:iso639-3Code}, \texttt{plx:iso639-1Code} \\
    \midrule
    \texttt{plx:LanguageVariety}
                          & \texttt{plx:languageVarietyOf}, \texttt{rdfs:label} \\
    \midrule
    \texttt{plx:Iso639-1Code} & \\
    \midrule
    \texttt{plx:Iso639-3Code} & \\
    \midrule
    \texttt{plx:Expression}
                          & \mbox{\texttt{plx:languageVariety}, \texttt{plx:degradedText},} \texttt{rdfs:label} \\
    \midrule
    \texttt{plx:Meaning}
                          & \mbox{\texttt{plx:source}, \texttt{plx:identifier},} \texttt{plx:meaningDefinition} \\
    \midrule
    \texttt{plx:Definition}
                          & \texttt{plx:languageVariety}, \texttt{rdfs:label} \\
    \midrule
    \texttt{plx:Denotation}
                          & \texttt{plx:denotationMeaning}, \texttt{plx:denotationExpression}, \texttt{plx:wordClass} \\
    \midrule
    \texttt{plx:WordClass}  
                          & \texttt{rdfs:label} \\
    \midrule
    \texttt{plx:License}  & \texttt{rdfs:label} \\
    \bottomrule
  \end{tabular}
  \end{tiny}
  \caption{Classes and properties used in the PanLex RDF vocabulary. Note that all \texttt{rdf:type} properties are omitted for brevity.}
  \label{tbl:vocabulary}
\end{table}
